# One censored outcome

```{r setup}
# Setup code
library(cmdstanr)
# ggplot2 theme setup
library(ggplot2)
ggplot2::theme_set(hgp::theme_ms())
pth_base <- here::here("examples", "simple-censored-outcome")
```

Now that we've gone through a very broad overview of what censored data is,
we'll discuss a simple example with a censored outcome.

For this example, we'll stay on the theme of concentrations that we started
in the last chapter. Analyzing immunological, medical, or environmental
concentrations is a common use case of censored data that doesn't require us
to also discuss the intricacies of time-to-event analysis. This time,
we'll consider **the amount of glyphosate in household drinking water.**

## Glyphosate data simulation

In this first example, since we're trying to understand these models, we'll
**simulate our data** from a known generative model. This allows us to be
confident that our model for censored data is actually helping us to recover
the correct parameters. Once we're more comfortable with the model, we can
try to analyze some data with unknown generative processeses.

[Glyphosate](https://en.wikipedia.org/wiki/Glyphosate) is an organophosphonate pesticide originally marketed as Roundup by Monsanto. Roundup was quickly
adopted for industrial agriculture, especially after the introduction of
genetically modified Roundup-resistant crop species in the mid-90's. Due to
widespread agricultural use in the US, glyphosate is an increasingly common
groundwater contaminant, with a [Maximum Containment Level Goal of 0.7 parts
per million in tap water set by the EPA](https://web.archive.org/web/20240509003411/https://www.epa.gov/sites/default/files/2015-06/documents/epa-200.7.pdf).

1. **Source population.** Our fictional source population will be $n$ randomly
sampled households from a fictional city, indexed from $i = 1, \ldots, n$. We
only have one measurement per household. For the purposes of our study,
we'll assume that each of these houses has their own water supply (which is
unrealistic but sufficient for a censored data tutorial).
1. **Outcome variable.** For this example, our outcome (or dependent) variable
is $y_i$, the log concentration of glyphosate in parts per million (ppm) detected
from the tap in household $i$ using an analytical test.
1. **Censoring structure.** Our investigators have decided to use [the test described in this paper](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6163928/), which has a lower limit of detection of 2 µM. Given a molar mass of 169.07 g/Mol
for glyphosate, we can convert this to an LoD of 0.33 ppm, just under half of
the MCLG. We assume there is no upper LoD or interval censoring.
1. **Independent variables.** For the sake of this example, we will measure a
few independent variables at the household level. Without doing any research
on the actual patterns of glyphosate distribution in groundwater, we'll assume
that glyphosate concentration is affected by:
	1. distance to the nearest agriculture site (km).
	1. whether the home uses glyphosate-based pesticides for personal gardening.
	1. whether the home has a water filter on the primary tap.

To get our data, we'll first randomly generate our covariates. Let's assume that
our town of interest is a square, so we can normalize all of the distances so
that each side of the square has length 1. Then (again somewhat unrealistically)
we'll assume that home X and Y coordinates are independently drawn from a
uniform distribution on $(0, 1)$. For convenience, we'll place the only farm
in town at $(0.5, 0.5)$ and calculate the distances. Then since the other
two independent variables are binary, we'll randomly sample those for each house,
say with respective probabilities $0.2$ and $0.4$.

```{r gly data independent variables}
set.seed(370)
N <- 147
gly_preds <-
	tibble::tibble(
		house_x_coord = runif(N, 0, 1),
		house_y_coord = runif(N, 0, 1),
		dist_from_farm = sqrt((house_x_coord - 0.5)^2 + (house_y_coord - 0.5)^2),
		personal_gly_use = rbinom(N, 1, 0.2),
		water_filter_use = rbinom(N, 1, 0.4)
	)
head(gly_preds)
```

Now we need to describe our data generating model for the outcome. Of course,
I am just going to randomly pick some parameters and mess around until they
look about right -- we know that our glyphosate levels should be somewhere in
the neighborhood of $(0, 1.4)$-ish. We'll use a linear model, which means
we assume that:
$$
\begin{aligned}
\log(y_i^*) &\sim \text{Normal}(\mu_i, \sigma^2) \\
\mu_i &= \beta_0 + \beta_1 x_{1, i} + \beta_2 x_{2, i} + \beta_3x_{3, i}.
\end{aligned}
$$
Here, $y_i$ is the concentration of glyphosate in tap water of house $i$,
$$x_{1, i}$$ is the distance from house $i$ to the farm, $x_{2, i}$ is the
personal glyphosate use variable, and $x_{3, i}$ is the water filter use
variable. So we need to pick all four of those $\beta_p$ coefficients and the
value of $\sigma^2$, the residual variance, before we can simulate our glyphosate
levels.

```{r simulating y values}
# Make a table of true coefs for using later
coefs <-
	tibble::tribble(
		~term, ~estimate,
		"(Intercept)", 0.8,
		"dist_from_farm", -4,
		"personal_gly_use", 0.5,
		"water_filter_use", -0.4
	)
res_sd <- 0.25

gly_data <-
	gly_preds |>
	dplyr::mutate(
		mu = 0.8 - 4 * dist_from_farm + 0.5 * personal_gly_use -
			0.4 * water_filter_use,
		y_star = exp(rnorm(N, mu, res_sd))
	)

LoD <- 0.33

summary(gly_data)
```

You can see from the above simulation code that the values I ended up choosing
are as follows:

$$
\begin{aligned}
\log(y_i^*) &\sim \text{Normal}(\mu_i, 0.5^2) \\
\mu_i &= -2.5 + 2\cdot x_{1, i} + 0.5\cdot x_{2, i} -0.4\cdot x_{3, i}.
\end{aligned}
$$
These values seemed to give a reasonable range of $y$ values (on the natural
scale), and have signs that made sense to me. The intercept represents
the concentration of glyphosate expected in tap water for a person who lives
on the farm, no personal glyphosate use, and no water filter, and is
about $2.23$ ppm, which is quite high and perhaps expected for the point source
of the contaminant. With these parameters, we will have
$`r round(mean(gly_data[['y_star']] <= 0.33) * 100, 1)`\%$ of data points below
the limit of detection, which is not ideal (of course the ideal is zero
percent), but not too bad either.

Our censoring model looks like this:

$$
y_i = \begin{cases}
L, & y_i^* < 0.33 \text{ ppm} \\
y_i^*, & \text{otherwise}
\end{cases}.
$$
Our censoring indicator $C_i$ will look like this:

$$
c_i = \begin{cases}
1, & y_i^* < 0.33 \text{ ppm} \\
0, & \text{otherwise}
\end{cases}.
$$
Let's first apply the censoring to our data. We'll arbitrarily choose $L = 0$.
Then we'll take a look at the data we would actually observe.

```{r}
# Dataset including latent variables
# Do the censoring
L <- 0.1
gly_data_lnt <-
	gly_data |>
	dplyr::mutate(
		# Create the censoring indicator
		c = ifelse(y_star <= LoD, 1, 0),
		# Create the censored outcome
		y = ifelse(c == 1, L, y_star)
	)

# Dataset including ONLY the observed variables
gly_data_obs <- gly_data_lnt |>
	dplyr::select(dist_from_farm, personal_gly_use, water_filter_use, c, y)

head(gly_data_obs)
```

Note that setting $y_i = 0.1$ if the value is censored is **completely arbitrary**.
Many people will set it to a value like the LoD, or half the LoD, or some
crazy thing with $\sqrt{2}$ in it, and then pretend those are the real values.
**All of these arbitrary values are equally bad.** Let's look at the
distribution of the latent and observed values just to show this.
In real life, we can't see this, but this example should remind us that picking
an arbitrary number is not very good.

```{r}
# Arrange the data correctly for plotting
gly_data_lnt |>
	dplyr::select(
		"observed" = y,
		"latent" = y_star
	) |>
	tidyr::pivot_longer(dplyr::everything()) |>
	# Now make the plot
	ggplot() +
	aes(x = value) +
	geom_histogram(
		boundary = 0,
		binwidth = 0.1,
		color = "black",
		fill = "gray"
	) +
	facet_wrap(vars(name)) +
	ggtitle(paste0("Censored values coded as ", L))
```

Let's also take a look at what the data would have looked like if we set, say
$L = 0.33$ (the LoD).

```{r}
gly_data_lnt |>
	# Set censored values to LoD
	dplyr::mutate(y = ifelse(y == L, 0.33, y)) |>
	# Arrange the data for the plot
	dplyr::select(
		"observed" = y,
		"latent" = y_star
	) |>
	tidyr::pivot_longer(dplyr::everything()) |>
	# Make the plot
	ggplot() +
	aes(x = value) +
	geom_histogram(
		boundary = 0,
		binwidth = 0.1,
		color = "black",
		fill = "gray"
	) +
	facet_wrap(vars(name)) +
	ggtitle("Censored values coded as LoD")
```

We can see how this histogram makes the censored values look like actual data
(cat screaming emoji)!! Whereas the previous set of histograms with a spike at
zero should signal that there is something strange going on in the data. So
substitution can cause data to be misleading. For this reason, it can sometimes
be useful for analysts to record values as "Nondetect" or "< LoD" in the dataset
(or some other kind of text indicating it is not a regular number),
forcing the analyst to clean up the data before it can be statistically examined.
The problem can be somewhat avoided if we include an explicit indicator of
censoring in our data, like so.

```{r}
gly_data_lnt |>
		# Set censored values to LoD
	dplyr::mutate(y = ifelse(y == L, 0.33, y)) |>
	dplyr::select(
		"observed" = y,
		"latent" = y_star,
		c
	) |>
	tidyr::pivot_longer(-c) |>
	# Now make the plot
	ggplot() +
	aes(x = value, fill = factor(c)) +
	geom_histogram(
		boundary = 0,
		binwidth = 0.1,
		color = "black",
		alpha = 0.5,
		position = "stack"
	) +
	facet_wrap(vars(name)) +
	scale_fill_manual(values = c("#E69F00", "#56B4E9"), name  = "Below LoD?") +
	ggtitle(paste0("Censored values coded as LoD"))
```

```{r}
gly_data_lnt |>
	dplyr::select(
		"observed" = y,
		"latent" = y_star,
		c
	) |>
	tidyr::pivot_longer(-c) |>
	# Now make the plot
	ggplot() +
	aes(x = value, fill = factor(c)) +
	geom_histogram(
		boundary = 0,
		binwidth = 0.1,
		color = "black",
		alpha = 0.5,
		position = "stack"
	) +
	facet_wrap(vars(name)) +
	scale_fill_manual(values = c("#E69F00", "#56B4E9"), name  = "Below LoD?") +
	ggtitle(paste0("Censored values coded as ", L))
```

Here, we can see that the histogram conflates part of the censored and
non-censored values because of the binwidth we set. All that is to show, when
there is a possibly of censored data, we should be extra careful as analysts to
make sure we aren't computing incorrect statistics.

As another instructive example, let us first attempt to estimate the mean and SD
of the glyphosate concentrations. We know an unbiased estimate of marginal mean
and CI (that is, the statistic if we ignore all of the `x` values), because we
have the underlying latent values. So let's estimate those first.
(Because we have a normal distribution, we could probably get the analytical
marginal mean assuming unknown $x_i, p$ values, but we won't do that here.)
In R, we can quickly construct the Wald-type CI based on the `t`-distribution
using the `t.test()` function.

```{r}
latent_t_test <-
	gly_data_lnt$y_star |>
	# Remember we made a log-normal assumption so we take the log here
	log() |>
	t.test() |>
	broom::tidy() |>
	# Re-exponentiate the results
	dplyr::mutate(dplyr::across(c(estimate, conf.low, conf.high), exp))
print(latent_t_test)
```

Now, if we compute the same test using the observed values, we can see what
happens.

```{r}
observed_t_test <-
	gly_data_obs$y |>
	# Remember we made a log-normal assumption so we take the log here
	log() |>
	t.test() |>
	broom::tidy() |>
	# Re-exponentiate the results
	dplyr::mutate(dplyr::across(c(estimate, conf.low, conf.high), exp))
print(observed_t_test)
```

The estimate is much lower! In fact, the 95% CI doesn't even cover the latent
estimate! Of course, **we can arbitrarily change the estimate by recoding the
censored values.** If we bumped them up to the LoD, the estimate would go up
and if we made them lower, it would go down.

Despite knowing this, let's see what happens in our linear model.

## Naive linear models

Now, if we were entrusted with a data set for analysis and had no idea it was
censored, we would typically assume that values at the LoD are measured
exactly. Though, as we discussed, some EDA might be suggestive, we will fit
the ordinary linear regression model we described in the data generating
process earlier (here, we are fortunate enough to know that this is an
appropriate model, which is always untrue in the real world).

```{r}
# Fit the linear model
naive_lm <-
	lm(
		log(y) ~ 1 + dist_from_farm + personal_gly_use + water_filter_use,
		data = gly_data_obs
	)
# Print the results in a little table
naive_lm |>
	broom::tidy(conf.int = TRUE) |>
	tibble::add_column(truth = coefs$estimate) |>
	dplyr::select(term, estimate, conf.low, conf.high, truth) |>
	knitr::kable(digits = 3)
```

Here in the table we can see that all of the estimates are biased away from the
null, which is exactly what we don't want in this kind of study -- we would, if
anything, prefer that they be biased towards the null so we avoid overstating
the effect. Importantly, we can see that the CI's for `dist_from_farm`
and `personal_gly_use` **do not even contain the true value**! So even though
we know the exact data generating process, and we know our model reflects that,
**if we don't account for censoring, we can get completely wrong estimates**!

So, then, what are we to do?

## Integration method for censored data

We can regain some measure of our lost diginity using the integration trick we
discussed in the introduction. Of course, in the introduction, we only talked
about adjusting for censoring in the univariate case, but fortunately we
are modeling the **conditional distribution of $y$** so we can use the same
trick:

$$
\begin{aligned}
\mathcal{L}\left(\theta \mid y_i, x_i \right) &= f_{Y_i \mid X_i = x_i}(y_i \mid \theta, x_i) \\
&= \bigg( f(y_i \mid \theta, x) \bigg)^{1 - c_i} \bigg( P(Y_i = y_i \mid x)\bigg)^{c_i} \\
&= \bigg( f(y_i \mid \theta, x_i) \bigg)^{1 - c_i} \bigg( \int_{-\infty}^{y_\min} f(y_i \mid \theta, x_i) \ dy_i \bigg)^{c_i} \\
&= \bigg( f(y_i \mid \theta, x_i) \bigg)^{1 - c_i} \bigg( F(y_\min \mid \theta, x_i) \bigg)^{c_i}.
\end{aligned}
$$

The likelihood for $y_i$ is easy to write out here since the censoring structure
is (relatively) simple. This gives rise to the likelihood of the sample,
which (under the assumption of mutual conditional independence) is
$$
\mathcal{L}\left(\theta \mid x, y \right) =  \prod_{i = 1}^n \mathcal{L}\left(\theta \mid y_i, x_i \right).
$$

Now that we've conducted the likelihood, we can do either of the usual things
we would do to estimate the parameters: find the argument $\theta$ that
maximizes the likelihood, or apply some priors and use an algorithm to estimate
a Bayesian posterior.

Doing either of these is not too complicated for this specific example -- we
can easily write a function to optimize, or we could do some kind of grid
or quadratic posterior approximation. For this example, neither of those is
very difficult and should converge easily. But, we have the benefit of
excellent statistical tools that have already been written, so we might as
well use them.

### Frequentist models

The `R` package `survival` (which actually predates the `R` language) implements
parametric models of this form for many common distributions. Don't let the
name fool you: we can do models other than survival analysis. In this particular
case, we just need to specify a parametric model with a normal distribution and
left censoring. The syntax for this is a bit strange, we need to use the
`Surv()` function to set up a "survival object" which we pass as the response
variable in the `survreg()` function.

```{r}
# First we have to transform the outcome
surv_model <- survival::survreg(
	# Creating the "survival time" outcome
	survival::Surv(
		# If the value is lower than LoD, replace it w/ LoD, then take the log
		pmax(y, LoD) |> log(),
		# The censoring indicator needs to be the opposite of what makes sense --
		# zero for censored, one for uncensored
		!c,
		# Specify left censoring
		type = 'left'
	) ~
		# All the other linear model stuff as usual
		dist_from_farm + personal_gly_use + water_filter_use,
	data = gly_data_obs,
	dist = "gaussian"
)

surv_model |>
	broom::tidy(conf.int = TRUE) |>
	tibble::add_column(truth = c(coefs$estimate, log(res_sd))) |>
	dplyr::select(term, estimate, conf.low, conf.high, truth) |>
	knitr::kable(digits = 3)
```

Despite all the finagling we have to do, we can see that the estimates are now
much better. Although the point estimates are not as close as we would expect
if we had the latent uncensored outcome variable, the confidence intervals
actually contain the true values this time. (Note that `Log(scale)` is the
estimated residual SD on the log scale, but for some reason the CI doesn't
get calculated by any of the `survival` methods.) Notably, the CIs for
this model are actually smaller than the the CIs for the naive model, even
though we're assuming there is extra uncertainty in the outcome. But we have
accounted for this uncertainty in the model, so it doesn't leak into the
parameter estimates (of course that's a non-technical explanation).

The worst part here is doing all that `Surv()` stuff, but fortunately there are
ways to avoid having to do all that. This method is commonly called the "Tobit
model" in econometrics, and there is a well-developed literature around this
model, and some variations and extensions. In particular, the `AER` package
provides a function called `tobit()` that allows one to specify these kind of
simple censoring models in standard `R` syntax, and automatically does the
conversion to a `survreg()` model.

```{r}
tobit_model <- AER::tobit(
	log(y) ~ dist_from_farm + personal_gly_use + water_filter_use,
	data = gly_data_obs,
	left = log(LoD),
	right = Inf,
	dist = "gaussian"
)

summary(tobit_model)
```

Now, if we want, it's also not too bad to do this in a Bayesian framework.

### Bayesian models

First we need to describe some basic priors that will work. For this example, we
know that the parameters actually have "true values" so assigning a degenerate
prior would actually reflect our true beliefs, but it's silly. So we'll assign
some usual weakly informative priors that will help our model work right. If you
want to do the priors with the variance of 10000 to be "objective" or whatever
you can do that but I don't think it's a good idea. I'll base my priors largely
on the advice in [Statistical Rethinking by Richard
McElreath](https://xcelab.net/rm/) and the [Stan Prior Choice
Guide](https://github.com/stan-dev/stan/wiki/Prior-Choice-Recommendations).

We need a prior for everything in our model that isn't observed in the data. For
this model, that's our residual variance and the beta coefficients. We'll
set the priors as follows:

$$
\begin{aligned}
\beta_j &\sim \mathrm{Normal}\left(0, 3\right), \quad j = {0, 1, 2, 3}; \\
\sigma &\sim \text{Half-t}\left(0, 3, \nu = 3\right).
\end{aligned}
$$
here, we use $\mathrm{t}\left(\mu, \sigma, \nu\right)$ to denote the
location-scale Student's $t$-dsitribution with $\nu$ degrees of freedom.
We use normal distributions centered at 0 for the $\beta_j$ to indicate
our skepticism about the strength of the effects (if we were to pretend we
didn't know the generative model). Because $\sigma$ must be postive, we
actually use the "half" version of the distribution, which has strictly
positive support.

Now, for this kind of model where only the outcome is censored, we can actually
have the super-handy `brms` R package do the heavy lifting for us. All we have
to do is a bit of data cleaning, and a bit of working specifying the model
structure, but the package will handle writing, compiling, and running all of
the Stan code, which is very conveninent. First we'll fit the naive model,
to show the `brms` syntax, and then we'll cover the censored adjustment.

#### naive model in `brms`

At first glance, the `brms` syntax appears to be quite difficult to use.
While complex, it is concise and specifying the syntax is much less difficult
than writing Stan code for many models. Fitting a `brms` model broadly
requires 5 things from us: the first two are the formula, in `lm()` type syntax
with a few extras, and the data, as we expect. We also need to specify a
distributional `family` argument for the likelihood, which we implicitly do
when we call `lm()`. As with base `R`, the `gaussian` likelihood is the default
in `brms`, but it's good practice to be clear. We also need to specify the
`priors`, which are written in a specific way, and the Stan control parameters
which control the Bayesian routine. Because `brms` is a wrapper for Stan's
NUTS sampler, the fitting routine is much more complicated than the routine for
`lm()` and there's a lot we can change.

If we just ignore these arguments, we can still get some results. Note that
when you run a `brms` model like this, you'll first get a compiling message,
then several intermittant progress updates about how fitting is going.
I have those silenced here because it generates a lot of them.

```{r}
#| message: false
library(brms)
basic_brms_fit <- brms::brm(
	formula = log(y) ~ 1 + dist_from_farm + personal_gly_use + water_filter_use,
	data = gly_data_obs
)
summary(basic_brms_fit)
```

Because the model is simple, we don't have to do much. The estimates here are
very similar to the estimates from the naive frequentist model.
Let's apply some priors and some
Stan arguments and see what happens. First, we'll set up the priors.

I recommend reading the `brms` documentation and papers to learn more about
how this prior syntax works, but we can also take a look at the default
priors that `brms` set for our basic model.

```{r}
prior_summary(basic_brms_fit)
```

Here, we can see that the coefficients have `(flat)` priors, which often works
for this simple of a model, but is usually not a good choice because it can
be difficult for the sampler. The default prior for `sigma`, the residual SD,
is very similar to what I picked, and I think is generally a good default
prior. Anyways, based on the `class` column from this output, we can figure
out how to set up a prior. Note that the `lb` for sigma is the *lower bound*.

```{r}
my_priors <- c(
	prior(normal(0, 3), class = "b"),
	prior(normal(0, 3), class = "Intercept"),
	prior(student_t(3, 0, 3), class = "sigma", lb = 0)
)

my_priors
```

Everything else we pass in, other than these four arguments, will be Stan
control parameters. If you want to learn about Stan specifics, there are way
better resources than this, so I recommend you read those. I just want to
make it clear that these other things are telling Stan how it should run the
fitting routine.

```{r}
#| message: false
naive_brms_fit <- brms::brm(
	formula = log(y) ~ dist_from_farm + personal_gly_use + water_filter_use,
	data = gly_data_obs,
	family = gaussian(),
	prior = my_priors,
	warmup = 1000,
	iter = 2000,
	chains = 4,
	cores = 4,
	seed = 32134,
	backend = "cmdstanr"
)
summary(naive_brms_fit)
```

The estimates don't change that much from the previous fit. Again, that's
because a simple linear model with data generated from a linear model is really
easy to handle. Once you start getting into more complicated models, the
priors and stan arguments can matter a lot. But now that we've got all
that taken care of, we should talk about censoring.

#### censored model in `brms`

In `brms`, we can handle censoring by a special formula syntax that tells the
model to do a likelihood adjustment. Notably, how to do this is kind of
buried in the documentation. And at time of writing, it never explains that the
likelihood adjustment is what it actually does (but it is). The best place to
read about this functionality and other additional options for `brms` is on
[this page](https://paul-buerkner.github.io/brms/reference/brmsformula.html) of
the documentation. We need to do a bit of data cleaning first before we can
handle the censoring though. As you can see on that page, we need our censoring
indicator variable to contain `-1` for left-censored data, instead of `1` like
we currently have. The censoring bound should be given in `y`, so we need to
transform it again so that all censored values are written down as the actual
LoD.

```{r}
brms_data <-
	gly_data_obs |>
	dplyr::mutate(
		# Transform censoring indicator
		censored = ifelse(c == 1, -1, 0),
		# Transform the outcome
		outcome = pmax(log(y), log(LoD)),
		# Only keep the variables we plan to give to brms
		.keep = "unused"
	)

head(brms_data)
```

Now we can fit the `brms` model with the censoring correction.

```{r}
#| message: false
cens_brms_fit <- brms::brm(
	formula = outcome | cens(censored) ~ dist_from_farm + personal_gly_use +
		water_filter_use,
	data = brms_data,
	family = gaussian(),
	prior = my_priors,
	warmup = 1000,
	iter = 2000,
	chains = 4,
	cores = 4,
	seed = 32134
)
summary(cens_brms_fit)
```

Again, our results are similar to the frequentist model. All of the true
values are now inside of the estimated 95% credible intervals, with the
one exception of `sigma`, the residual SD, which had a true value of 0.25 --
so we'll say it's inside the credible interval up to measurement/computer
precision, there's no way that small of a difference would practically matter.

Unfortunately, not every model can be handled with `brms`, even though it's
very flexible. For example, when we talk about censored predictor values, we
won't be able to use `brms` for the approach we'll use. Instead, we'll have to
write our own Stan code, which is even more flexible than `brms` (though it
lacks many of the convenience features).

#### Custom Stan code with `cmdstanr`

STAN CODE HERE.

## Conclusions

So, 

## Appendix: notes on modeling {.appendix .unnumbered}

Since this is a simple example focused on dealing with the censoring adjustment,
I decided not to get into the weeds about modeling anywhere else. There are a
few technical choices I made/didn't make, so I decided to write them down
here for posterity, without cluttering up the main text.

* You should definitely use `cmdstanr` for running Stan models. In general,
it stays up to date much better, is a bit easier to use, and is compatible with
almost everything `rstan` is (and the formats are interchangable). It can be
annoying to install, but it's worth the hassle.
* A lot of people like flat or very diffuse priors because they are "objective".
I dislike them for a number of reasons. First of all, they often reduce the
ability of the model to sample efficiently. Second, due to the inherent
researcher degrees of freedom in setting up the model structure and what data
are present, I don't think talking about the "objectivity" of priors makes any
sense. Finally, the priors represent our prior beliefs about the parameters, and
a flat prior means we think any possible value is equally likely. This is why
people often think this is an "objective" prior but I think that's silly. I
certainly don't think my beta coefficient is equally likely to be 1 vs. negative
three billion and seven, but that's what the flat prior says. In general, I
know the effect should be fairly weak, so we should use skeptical priors that
allow the parameter to get larger if the data support that.
* In general for "generic priors", I tend to prefer $t$ and half-$t$ priors
with 3 degrees of freedom. These have thicker tails than normal/half-normal
priors, but not as thick as Cauchy priors, which some people will recommend.
A Cauchy prior reflects a belief that the variance of a variable is infinite,
which is not plausible to me, and Cauchy priors also cause sampling efficiency
to plummet in many cases without much of a benefit. So in general if I had to
pick a "generic default prior" it would be $t$ with three degrees of freedom.
* I didn't talk about centering and scaling here, but in general you should
typically scale your variables (e.g. divide by the MAD) before putting them
into Stan, simply because it helps the sampler.
* Centering can also be helpful, and if you want to do any kind of variable
selection or comparison between variables, you need to center them. Centering
can make coefficient interpretations more difficult, but one of the benefits of
Bayesian modeling is that any posterior calculation we want to make comes with
an entire posterior distribution of estimates, so we never have to worry about
figuring out CIs for a posteriori calculations. In general it's often better to
do these kind of computational tricks to make the model fit efficiently, and
then do transformations and predictions of interest afterwards.

## Appendix: imputation methods {.appendix .unnumbered}



<!-- END OF FILE -->
